Simple repression draft

### 2020-05-20 doc/section_02_means.tex
RP 9am Wed May 19
Finished going through section 2 on means.

Closing everything on my desktop until further notice.

\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
What have we learned from our exploration of the models
in~\fig{fig1:means_cartoons} and their predictions of fold-changes in mean
expression levels? The takeaway is the existence of a master curve
(\fig{fig1:means_cartoons}C) to which the fold-change predictions of all the
models collapse. This master curve is parametrized by only two effective
parameters: $\Delta F_R$, which characterizes the repressor, and $\rho$, which
characterizes all other features of the promoter architecture. The key
assumption underpinning this result is that zero transcription occurs when a
repressor is bound to its operator. Note, however, that we are agnostic about
the molecular mechanism which achieves this; steric effects are the obvious
mechanism, but, for instance, ``action-at-a-distance'' mediated by kinked DNA
due to repressors bound tens or hundreds of nucleotides upstream of a promoter
is plausible as well.

Why does the master curve of~\fig{fig1:means_cartoons}C exist at all? This
brings to mind the deep questions posed in, e.g.,~\cite{Frank2013}
and~\cite{Frank2014a}, asking us to search for general ``symmetries'' of the
problem instead of specifying generative models of the known microscopics of a
system. In our case, the key feature seems to be the exclusive nature of
repressor and RNAP binding, which allows the parameter describing the repressor,
$\Delta F_R$, to cleanly separate from all other details of the promoter
architecture, which are encapsulated in $\rho$. In the graph-theoretic language
which was applied to similar problems in~\cite{Gunawardena2012}
and~\cite{Ahsendorf2014}, the repressor-bound state is a leaf of the graph
describing transitions between promoter states. Arbitrary nonequilibrium
behavior can occur on the rest of the promoter state space, but it may all be
swept up in the effective parameter $\rho$, to which the repressor makes no
contribution.

@@

How do we test this and probe the theory with fold-change measurements? There is
a fundamental limitation in that the master curve is essentially a one-parameter
function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the value of $\rho$,
such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
and/or, RNAP clusters analogous to those obvserved in
eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
precision, inferring the potential microscopic contributions to $\rho$, buried
inside a log no less, from fold-change measurements seems hopeless. As a
statistical inference problem it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

@@

One might instead imagine a bottom-up modeling approach, where we mathematize a
model of what we hypothesize the important steps are and are not, use in vitro
data for the steps deemed important, and \textit{predict} what $\rho$ should be.
But again, because of the one-parameter nature of the master curve, many
different models will likely make indistinguishable predictions, and without any
way to experimentally perturb in vivo, there is no clear way to test whether the
modeling assumptions are correct.

@@

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}. In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
meaning therefore $\rho\approx1$. This is suggestive, but how can we be sure
that $\rho$ is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a
compensatory factor? A resolution is offered by an independent inference of
$\rho$ in the absence of repressors. This was done in~\cite{Razo-Mejia2020} by
fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons}, with zero
repressor (looking ahead, this is equivalent to model 4
in~\fig{fig:constit_cartoons}), to FISH data from~\cite{Brewster2014}. This
provided a determination of $k^+$ and $k^-$, from which their ratio is estimated
to be no more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

@@

Moving forward, we have weak evidence supporting the interpretation of $\Delta
F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
discriminatory power to test the theory and distinguish the various models with
measurements of fold-changes in means, how do we probe the theory? Clearly to
discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
need to go beyond means to ask questions about kinetics, noise and even full
distributions of mRNA copy numbers over a population of cells. If the
``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
free energy of repressor binding inferred from fold-change measurements should
agree with repressor binding and unbinding rates. In other words, the equilbrium
and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
\begin{equation}
\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
        = - \log(k_R^+/k_R^-),
\end{equation}
must hold, where $\beta\Delta\epsilon_R$ is inferred from the master curve fit
to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
have directly observed these rates, and in the remainder of this work we explore
a complementary approach: inferring repressor binding and unbinding rates
$k_R^+$ and $k_R^-$ from FISH measurements of mRNA population distributions.

\mmnote{Add a notation aside somewhere that all association rates are written as zeroth-order rates. In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}

>>>>

\subsubsection{Summary of Results Ccross models for Fold-Changes in Mean Expression}
What have we learned from our exploration of the models
in~\fig{fig1:means_cartoons} and their predictions of fold-changes in mean
expression levels? The key outcome of our analysis is the existence of a master curve
 shown in \fig{fig1:means_cartoons}(C) to which the fold-change predictions of all the
models collapse. This master curve is parametrized by only two effective
parameters: $\Delta F_R$, which characterizes the number of repressors and
their binding strength to the DNA, and $\rho$, which
characterizes all other features of the promoter architecture. The key
assumption underpinning this result is that no transcription occurs when a
repressor is bound to its operator. Note, however, that we are agnostic about
the molecular mechanism which achieves this; steric effects are one plausible
mechanism, but, for instance, ``action-at-a-distance'' mediated by kinked DNA
due to repressors bound tens or hundreds of nucleotides upstream of a promoter
is plausible as well.

Why does the master curve of~\fig{fig1:means_cartoons}(C) exist at all? This
brings to mind the deep questions posed in, e.g.,~\cite{Frank2013}
and~\cite{Frank2014a}, asking us to search for general ``symmetries'' of the
problem instead of specifying generative models of the known microscopics of a
system. In our case, the key feature seems to be the exclusive nature of
repressor and RNAP binding, which allows the parameter describing the repressor,
$\Delta F_R$, to cleanly separate from all other details of the promoter
architecture, which are encapsulated in $\rho$. In the graph-theoretic language
which was applied to similar problems in~\cite{Gunawardena2012}
and~\cite{Ahsendorf2014}, the repressor-bound state is a leaf of the graph
describing transitions between promoter states. Arbitrary nonequilibrium
behavior can occur on the rest of the promoter state space, but it may all be
swept up in the effective parameter $\rho$, to which the repressor makes no
contribution.

@@

How do we test this and probe the theory with fold-change measurements? There is
a fundamental limitation in that the master curve is essentially a one-parameter
function of $\Delta F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the value of $\rho$,
such as RNAP binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015},
and/or supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
and/or, RNAP clusters analogous to those obvserved in
eukaryotes~\cite{Cisse2013, Cho2016} and recently also observed in
bacteria~\cite{Ladouceur2020}. Even if $\Delta F_R$ is measured to high
precision, inferring the potential microscopic contributions to $\rho$, buried
inside a log no less, from fold-change measurements seems beyond reach. As a
statistical inference problem it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

@@

One might instead imagine a bottom-up modeling approach, where we mathematicize a
model of what we hypothesize the important steps are and are not, use in vitro
data for the steps deemed important, and \textit{predict} what $\rho$ should be.
But again, because of the one-parameter nature of the master curve, many
different models will likely make indistinguishable predictions, and without any
way to experimentally perturb in vivo, there is no clear way to test whether the
modeling assumptions are correct.

@@

We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
is essentially a one-parameter model, the one parameter being $\Delta F_R +
\log\rho$. By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
on the internal structure of $\rho$ itself. A hint is provided by the weak
promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
model in~\fig{fig1:means_cartoons}. In that picture, the weak promoter
approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.  This approximation can be well
justified on the basis of the number of RNAP per cell and the strength of
binding of RNAP to DNA at weak promoters.  This is suggestive, but how can we be sure
that $\rho$ is not, for instance, actually $10^2$ and that $\Delta F_R$ hides a
compensatory factor? A resolution is offered by an independent inference of
$\rho$ in the absence of repressors. This was done in~\cite{Razo-Mejia2020} by
fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons}, with zero
repressor (looking ahead, this is equivalent to model 4
in~\fig{fig:constit_cartoons}), to FISH data from~\cite{Brewster2014}. This
provided a determination of $k^+$ and $k^-$, from which their ratio is estimated
to be no more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

@@

Moving forward, we have weak evidence supporting the interpretation of $\Delta
F_R$ as a physically real free energy~\cite{Landman2019} and other work casting
doubt~\cite{Hammar2014}. How might we resolve the confusion? If there is no
discriminatory power to test the theory and distinguish the various models with
measurements of fold-changes in means, how do we probe the theory? Clearly to
discriminate between the nonequilibrium models in~\fig{fig1:means_cartoons}, we
need to go beyond means to ask questions about kinetics, noise and even full
distributions of mRNA copy numbers over a population of cells. If the
``one-curve-to-rule-them-all'' is more than a mathematical tautology, then the
free energy of repressor binding inferred from fold-change measurements should
agree with repressor binding and unbinding rates. In other words, the equilbrium
and nonequilibrium definitions of $\Delta F_R$ must agree, meaning
\begin{equation}
\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
        = - \log(k_R^+/k_R^-),
\end{equation}
must hold, where $\beta\Delta\epsilon_R$ is inferred from the master curve fit
to fold-change measurements, and $k_R^+$ and $k_R^-$ are inferred in some
orthogonal manner. Single molecule measurements such as from~\cite{Hammar2014}
have directly observed these rates, and in the remainder of this work we explore
a complementary approach: inferring repressor binding and unbinding rates
$k_R^+$ and $k_R^-$ from FISH measurements of mRNA population distributions.
\marginpar{\it MM: Add a notation aside somewhere that all association rates are written as zeroth-order rates. In other words, the concentration of the molecule is hidden inside, so for instance, $k_R^+$ is proportional to repressor copy number.}
